Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment

### 2016-06-24 removals/manuscript/removal_ms.tex
edits on the ms

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function (Chu and Adler 2015), but here we use a data-driven approach (Teller et al. 2016). We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $j$ on species $m$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, which means that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input coefficients, 
calls the model-fitting function \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
survival and growth. In addition, 
for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}. }
\comment{SPE: see Robin's comment about whether you mean $F_{mm}$ or $F_{jj}$; I'm not sure which one you did.}

@@

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{ \left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
\label{eqn:recrProcessModel}
\end{equation}
where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location,
$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term. We previously used a Ricker equation (Adler et al. 2010), but using the square-root transformation gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al. 2002) for the recruitment model. Following previous work (Adler et al. 2010), we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.

Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in which the quadrat is located:
$C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$								
where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.

@@

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose area (on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The density function for size $v$ at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))n(u_{j,t})   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how $\boldsymbol{\bar{W}}_j$ is calculated). The integral is evaluated over all possible sizes from a lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models:
\begin{equation}
k_j (v_j,u_j,\boldsymbol{\bar{W}}_j ) = S_j(u_j,\boldsymbol{\bar{W}}_j  (u_j)) G_j(v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j)) 
+ R_j (v_j,u_j,\boldsymbol{\bar{W}}_j) .
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}). The recruitment rate, $\lambda$, 
defined in (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}), gives the expected 
number of new recruits produced per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, 
we assumed that individual fecundity increases linearly with individual size (area, equal to $e^u$). The recruitment kernel is thus
$R_j(v_j,u_j,\boldsymbol{\bar{W}}_j)=C_0(v_j)e^{u_j}\lambda_j$ (Adler et al. 2010) where $C_0$ is the
size distribution of new recruits, and $\lambda_j$ is calculated from (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}) with
population densities converted into total cover per m$^2$, $C$.

In the vital rate regressions and in the IBM, the neighborhood crowding index, $W_ijm$ , unique to each individual $i$ based on the spatial locations and sizes of neighboring plants. This measure cannot be extended to the IPM, because the IPM is spatially implicit. Therefore, we used a spatially
implicit mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al. 2010). We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
but conspecific individuals displayed non-random, size-dependent spatial patterns: while small genets were randomly distributed, large genets 
were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM with a 'no-overlap' rule (Adler et al. 2010, 2012). In \ref{suppMethods} we explain our implementation of the no-overlap rule with our nonparametric competition kernels.

We simulated the IPM to estimate equilibrium abundances under three different scenarios. The first scenario featured all four species and our "baseline" model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models ("kernel selection" in the terminology of Metcalf et al. 20xx))). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. By comparing the output of this simulation with the previous scenario we can quantify the long-term consequences of differences between our baseline model and a model that includes additional removal treatment effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} (Fig. \ref{fig:CoverTrends}) , but did appear to cause an increase in the cover of \textit{P. spicata}. We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). Grass removal had no significant effect on \textit{A. tripartita} (95\% confidence interval on removal effect: -0.33, 0.27) or \textit{Poa secunda} (95\% CI: 0.02, 0.87) population growth but did increase \textit{H. comata} (95\% CI: -0.40, 0.28) and \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth (Table \ref{table:coefficients}). However, these treatment comparisons do not address our research questions about whether responses to removal were adequately predicted by models fit to observational data (i.e. control plots).

@@

To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilium cover. We concentrated on grass species' responses to \textit{A. tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation. Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A. tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig. \ref{fig:IPMresults}). Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H. comata}, but a 2.75 fold increase in the cover of \textit{P. spicata} (Fig. \ref{fig:IPMresults}).

@@

The small competitive release of many species in our system implies that classic resource competition is not the primary factor determining community structure. Interactions with above- or belowground enemies can also regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics. Species-specific responses to spatial and temporal environmental variation can also weaken interspecific interactions (Chesson et al. 2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al. 2009). More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. refs...). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.

@@

Our results have given us more confidence that competitive release is quite small for some species in some communities. But for many species in other communities, interspecific competition may still be an important driver. In closed canopy-forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al. 2010, Johnson et al. 2012). We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway (examples?). Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.  Determining where interspecific interactions and competitive release are strong, and where they are not, remains a fascinating research problem.

@@

Funding: NSF DEB-1353078 and DEB-1054040, Utah Agricultural Experiment Station (get journal paper number). The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.

@@

%%%%%%%%%%%%%%%%% For word count

@@

We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables. Because some $W=0$, we conducted two separate regressions. First, using all $W$'s, we performed a generalized linear regression with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s. Second, for the set of records in which the focal species $W>0$, we performed a linear regression. We repeated these regressions for individuals of each focal speices. Due to large samples size, $W$ values were often statistically significant predictors. However, they explained very little variance. The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r})

Spatially-implicit approximations of local crowding for the IPM. Steve? Paste in the MeanFieldSplineIPM.tex?

\section{Additional Tables}

>>>>

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function (Chu and Adler 2015), but here we use a data-driven approach (Teller et al. 2016). We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\label{eqn:wijm}
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}. }
\comment{SPE: see Robin's comment about whether you mean $F_{mm}$ or $F_{jj}$; I'm not sure which one you did.}

@@

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
\label{eqn:recrProcessModel}
\end{equation}
where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location,
$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term. We previously used a Ricker equation (Adler et al. 2010), but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al. 2002) for the recruitment model. Following previous work (Adler et al. 2010), we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.

Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in 
which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.

@@

\comment{SPE: how were group effects incorporated into the IPM, and the IBM?}

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose log area is between $u$ and $u+du$, hence their area 
(on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The size distribution function at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$ and $\vec{n}$ 
is the set of size-distribution functions for all species in the community. The integral is evaluated over all possible sizes from a 
lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al. 2010). 
We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
with a 'no-overlap' rule (Adler et al. 2010, 2012). In \ref{suppMethods} we explain how the spline competition kernels with 
the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.

For recruitment in the IPM, the factor $ \Phi_j  = \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'_{q,t}}} \right \rangle \right) }$
in eqn. (\ref{eqn:recrDataModel}) gives the expected 
area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, 
we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ (Adler et al. 2010) where $c_{0,j}$ is the
size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values
computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.

We simulated the IPM to estimate equilibrium abundances under three different scenarios. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of Metcalf et al. 2016))). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. By comparing the output of this simulation with the previous scenario we can quantify the long-term consequences of differences between our baseline model and a model that includes additional removal treatment effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but \new{shrub removal} did appear to cause an increase in the cover of \textit{P. spicata}. We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). \new{Removal treatment} had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.33, 0.27) or 
\textit{Poa secunda} (95\% CI: -0.40, 0.28) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.02, 0.87) and
 \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots. 
\comment{SPE: I changed the CI numbers in this paragraph to match what I think the Table means.}

@@

To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover. We concentrated on grass species' responses to \textit{A. tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation. Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A. tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig. \ref{fig:IPMresults}). Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H. comata}, but a 2.75 fold increase in the cover of \textit{P. spicata} (Fig. \ref{fig:IPMresults}).

@@

The small competitive release of many species in our system implies that classic resource competition 
is not the primary factor determining community structure. 
\comment{SPE: I think this depends on what you mean by ``classic resource competition''. If you mean Tilman-style competition for one limiting resource distributed homogeneously, then I agree. But a MacArthur-Levins model with resource spectrum and distinct utilization curves could also be called classic resource competition,
and that model will have weak to nonexistent competitive release if niche overlap is low.} 
Interactions with above- or belowground enemies can also regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics. Species-specific responses to spatial and temporal environmental variation can also weaken interspecific interactions (Chesson et al. 2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al. 2009). More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. refs...). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.

@@

Our results have given us more confidence that competitive release is quite small for some species in some communities. But for many species in other communities, interspecific competition may still be an important driver. In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al. 2010, Johnson et al. 2012). We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway (examples?). Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak. Determining where interspecific interactions and competitive release are strong, and where they are not, remains a fascinating research problem. \comment{SPE: Equally interesting, I think, is the question of why interspecific
interactions are so weak (when they are), as opposed to what you would expect from ``all plant species compete for water, light, and a few mineral resources''.}

@@

Funding: NSF DEB-1353078, DEB-1054040, DEB 1353039 (SPE), Utah Agricultural Experiment Station (get journal paper number). The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.

@@

@@

\subsection{Interspecific covariance in local crowding} 
We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables. Because some $W=0$, we conducted two separate regressions. First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' probability of having $W=0$ is influenced by the value of other species' $W$'s. Second, for the set of records in which the focal species has $W>0$, we performed a linear regression. We repeated these regressions for individuals of each focal species. Due to large samples size, $W$ values were often statistically significant predictors. However, they explained very little variance. The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}) \comment{It's not clear what the independent and dependent variables are, in the analyses described
here. It might help to use the notation from the Methods section?}

\subsection{Mean field approximation of local crowding for the IPM} 
\citet{adler-etal-2010} developed a mean field approximation for local crowding when the
competition kernels are all Gaussian functions, $F_{jm}(d) = e^{-\alpha_{jm} d^2}$. The approximation is explained in 
the online SI to \citet{adler-etal-2010} and in section 5.3 of \citet{ipmbook-2016}. 
Here we explain how that approximation was modified for the IPMs in this paper, which
used fitted nonparametric competition kernels.

For $j \ne m$ (between-species competition), overlap between individuals is allowed. The mean field approximation is 
that from the perspective of any focal plant in species $j$, individuals of species $m$ are distributed at random in space, 
independent of each other and of their size.

Consider the region between the circles of radius $x$ and $x+dx$ centered on a focal genet of species $j$. The area of this annulus
is $2 \pi x \; dx$  to leading order for $dx \approx 0$. A species $m$ genet 
in the annulus puts competitive pressure $F_{jm}(x)$ times its area on
the focal genet. The expected total competitive pressure from all such genets 
is therefore is $F_{jm}(x) 2 \pi x \; dx$ times the expected fractional cover of species $m$ in the annulus 
(fractional cover is the total area of species $m$ genets, as a fraction of the total area). The excepted fractional cover $C_m$ of species $m$
in the annulus equals its fractional cover in the habitat as a whole, because of the assumption of random distribution
spatial distributions. We therefore have $C_m  = \int e^u n_m(u,t) du/A$ where $A$ is the total area of the habitat. 
The total expected competitive pressure on a species-$j$ genet due to species $m$ is then 
\begin{equation}
W_{jm} = \int_0^\infty{C_m F_{jm}(x) 2 \pi x \; dx}  = C_m \left [2 \pi \int_0^{\infty} x F(x) \, dx \right ].
\label{eqn:wbarm}
\end{equation} 
The quantity in square brackets is a constant (that is, it only depends on what the kernel function
is) so it can be computed once and for all for each kernel used in the IPM. The integral is finite because
all fitted kernels fall to zero at a finite distance from the focal plant.

Our kernel fitting method only uses competition kernel values at the ``mid-ring'' distances
halfway between the inner and outer radii of a series of annuli around each focal
plant, scaled so that the value at the innermost mid-ring distance equals 1. 
In the IPM we defined the kernel at other distances by linear interpolation between values at 
mid-ring distances, except that for the innermost ring a kernel value of 1 was specified at the
outer radius of the ring and at distance $x=0$.

Now consider within-species competition. We assume that conspecifics cannot overlap. Genet shapes are irregular, but we 
nonetheless implement the no-overlap rule by assuming that a genet of log area $u_i$ is a 
circle of radius $r_i$ where $\pi r_i^2 = e^{u_i}$. The no-overlap rule is then that the centroids of two conspecific individuals 
must be separated by at least the sum of their radii.

For any one focal genet, the no-overlap restriction on its neighbors' locations affects 
only a negligibly small part of the habitat. The expected cover of individuals in the places
where they can occur (relative to one focal plant) is thus assumed to equal their expected locations
in the habitat as a whole.

Let $C_m(u)$ be the total cover of species $m$ genets of radius $r$ or smaller, 
\begin{equation}
C_m(r) = \int_L^{\log(\pi r^2)}{\! \! \! e^z n_m(z,t) \, dz} .
\label{eqn:cm}
\end{equation}
A focal genet of radius $r$ cannot have any conspecific neighbors centered 
at distances less than $r$. It can have a neighbor centered at distance $x>r$ if that neighbor's
radius is no more than $x-r$. Adding up the expected cover of all such possible neighbors
for a focal genet of radius $r$,    
\begin{equation}
W_{mm}(r) = 2 \pi \int_r^{\infty}F_{mm}(x) x C_m(x-r) \, dx
\label{eqn:wbarmr} 
\end{equation}
This integral is again finite and computable because the kernels $F$ fall to 0 at finite $x$.

\clearpage 
\newpage  
\section{Additional Tables}
